---
title: "Thesis Script"
author: "Kristen I. Hardy"
date: "2025-03-23"
output:
  pdf_document: default
  html_document: default
---

```{r setup, include=FALSE} 
knitr::opts_chunk$set(warning = FALSE, message = FALSE) 
```

## **Reading in the data files:**

## Metadata:
```{r}
Pipeline_metadata <- read.csv("GSC-2581_metadata.csv", header = TRUE, sep = ",")
```
 
## ID-matched dataframe:
```{r}
GSC_to_Carlsten <- read.csv("GSC-2581_fastq.csv", header = TRUE, sep = ",")
```
  
## GeneIDs and SampleID file:
```{r}
Gene_ID_and_Samples <- read.table(file = "salmon.merged.gene_counts_length_scaled.tsv", sep = '\t', header = TRUE)
```
    
## **Loading Packages:**
```{r, }
library(tidyr)
library(tibble)
library(dplyr)
library(BiocManager)
library(DESeq2)
library(RColorBrewer)
library(ggplot2)
library(ggrepel)
library(ggvenn)
library(devtools)
library(kableExtra)
```
    
## **Data Rangling:**
```{r}
Exposure_condition_column <- select(Pipeline_metadata,Exposure)

GSC_to_Carlsten_with_condition <- cbind(GSC_to_Carlsten, Exposure_condition_column)

Gene_ID_and_Samples_for_Object <-  read.table(file = "salmon.merged.gene_counts_length_scaled.tsv", sep = '\t', header = TRUE, row.names = 1)
  
count_data <- subset(Gene_ID_and_Samples_for_Object, select = -c(gene_name))
```

## Merge FA(DE) and FA(WS) gene counts for each subject --> accounts for FA differences:
```{r}
count_data <- count_data %>% 
  mutate(F145701_03 = F145701 + F145703, 
         F145706_08 = F145706 + F145708,
         F145711_13 = F145711 + F145713, 
         F145716_18 = F145716 + F145718, 
         F145721_23 = F145721 + F145723, 
         F145726_28 = F145726 + F145728, 
         .keep = "unused")
```

## More cleaning up:
```{r}
rounded_count_data <- round(count_data)

col_data <- select(GSC_to_Carlsten_with_condition, c(library.name, Exposure))

sample_patient_id_df = GSC_to_Carlsten_with_condition %>%
  select(library.name,library.external_identifier) %>%
  separate(library.external_identifier,into=c("patient_id","index_number"),sep="-")

col_data_new = col_data %>%
  left_join(.,sample_patient_id_df,by="library.name") %>%
  select(-index_number)
```

```

## Remove redundancies, call it "conditions" and then rearrange to have IC last:
```{r}
conditions <- unique(GSC_to_Carlsten_with_condition$Exposure)
conditions <- rev(conditions)
```

## **DESeq2:**

## Making the object:
```{r}
dds <- DESeqDataSetFromMatrix(countData = rounded_count_data, colData = col_data_new, design = ~ Exposure + patient_id)
```

## Run the differential expression analysis:
```{r}
dds <- DESeq(dds)
res <- results(dds)
```

## Pre-filtering to get rid of low count genes (did minimum 10 gene reads):
```{r}
smallestGroupSize <- 6
keep <- rowSums(counts(dds) >= 10) >= smallestGroupSize
dds <- dds[keep,]
```

## Set the reference for the contrasting results (DE vs DE(FA), etc):
```{r}
DE_results <- results(dds, contrast = c("Exposure", "DE", "FA"))
WS_results <- results(dds, contrast = c("Exposure", "WS", "FA"))
```

## Identifying significant padj values for the different conditions:

###DE vs FA:
```{r}
significant_DE <- DE_results[!is.na(DE_results$padj) & DE_results$padj < 0.05, ]
```

### WS vs FA:
```{r}
significant_WS <- WS_results[!is.na(WS_results$padj) & WS_results$padj < 0.05, ]
```

**# START OF DATA ANALYSIS:**

## Variance stabilizing transformation:
```{r}
vsd <- vst(dds, blind = FALSE)
```

## Plotting: Keep all the conditions consistently coloured:
```{r}
color_palette <- c(
  "FA" = "#d1b1fd",
  "IC"  = "#88e4ff",
  "WS" = "#fbb8ff",
  "DE" = "#f9e823")
```

## PCA plot of all conditions, to show that patients are grouped with themselves
### NOTE: renaming study participants:
```{r}
PCA_plotted_all <- plotPCA(vsd, intgroup = c("Exposure", "patient_id"), returnData=TRUE)
percentVar <- round(100 * attr(PCA_plotted_all, "percentVar"))
PCA_plotted_all2 <- PCA_plotted_all 
```

## Plot it:
```{r}
finalpcaplot_all <- ggplot(PCA_plotted_all2, aes(PC1, PC2, color = Exposure, shape = patient_id)) +
  geom_point(size=4) +
  scale_shape_manual(values = c(11, 13, 15, 16, 17, 18)) +
  xlab(paste0("PC1: ",percentVar[1],"% variance")) +
  ylab(paste0("PC2: ",percentVar[2],"% variance")) + 
  coord_fixed() +
  theme_minimal() +
  theme(
    panel.border = element_blank(),
    panel.grid.major = element_blank(),
    panel.grid.minor = element_blank(),
    axis.line = element_line(color = "black")
  ) +
  scale_color_manual(values=color_palette)
```

```{r, echo=FALSE}
  plot(finalpcaplot_all)
```

## **A little more data rangling:**

## DE first:
```{r}
DE_results_df <- as.data.frame(DE_results)
DE_results_df <- tibble::rownames_to_column(DE_results_df, "Gene_label")
```

## Join the "gene_id" column to the DE_results_df:
```{r}
DE_results_ensembldb = DE_results_df %>%
  left_join(.,Gene_ID_and_Samples, by = join_by(Gene_label == gene_id))
  DE_results_ensembldb <- subset(DE_results_ensembldb, select = -c(F145700, F145701, F145703, F145702, F145704,
                                                                    F145705, F145706, F145708, F145707, F145709,
                                                                   F145710, F145711, F145713, F145712, F145714,
                                                                   F145715, F145716, F145718, F145717, F145719,
                                                                   F145720, F145721, F145723, F145722, F145724,
                                                                   F145725, F145726, F145728, F145727, F145729))
```
  
## Do the same for WS:
```{r}
WS_results_df <- as.data.frame(WS_results)
WS_results_df <- tibble::rownames_to_column(WS_results_df, "Gene_label")
```
  
## Join the "gene_id" column to the WS_results_df:
```{r}
WS_results_ensembldb = WS_results_df %>%
    left_join(.,Gene_ID_and_Samples, by = join_by(Gene_label == gene_id))
  WS_results_ensembldb <- subset(WS_results_ensembldb, select = -c(F145700, F145701, F145703, F145702, F145704,
                                                                    F145705, F145706, F145708, F145707, F145709,
                                                                   F145710, F145711, F145713, F145712, F145714,
                                                                   F145715, F145716, F145718, F145717, F145719,
                                                                   F145720, F145721, F145723, F145722, F145724,
                                                                   F145725, F145726, F145728, F145727, F145729))
```

  
## **Aim 1: Volcano plots:**
    
##DE results volcano object:
```{r}
res_DE_volcanoFULL <- as.data.frame(DE_results_ensembldb)
  res_DE_volcanoFULL <- res_DE_volcanoFULL %>%
   dplyr::filter(!is.na(log2FoldChange) & !is.na(padj))  
    res_DE_volcanoFULL$Significant <- "NOT SIGNIFICANT"  
### if log2Foldchange > 0.5 and padjvalue < 0.05, set as "UP" 
  res_DE_volcanoFULL$Significant[res_DE_volcanoFULL$log2FoldChange > 0.5 & res_DE_volcanoFULL$padj < 0.05] <- "UP"
  
### if log2Foldchange < -0.5 and pvalue < 0.05, set as "DOWN"
  res_DE_volcanoFULL$Significant[res_DE_volcanoFULL$log2FoldChange < -0.5 & res_DE_volcanoFULL$padj < 0.05] <- "DOWN"
```
  
## Plot it:
```{r}
DEvolcanoFULL <- ggplot(res_DE_volcanoFULL, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
     scale_colour_manual(values = c("#ff5c68", "black", "#4ada51")) +
    geom_point(alpha = 0.6, size = 1.5) +
     geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") + 
     geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "blue") +
     geom_text_repel(data = subset(res_DE_volcanoFULL, Significant == c("UP", "DOWN")), max.overlaps = 18, aes(label = gene_name), vjust = 1.5, size = 3) +
    theme_minimal() +
    coord_fixed() +
    labs(x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")
  theme(
    panel.border = element_blank(),
    axis.line = element_line(color = "black")) 
```
```{r, echo=FALSE}
    plot(DEvolcanoFULL)
```
  
## WS results volcano object:
```{r}
res_WS_volcanoFULL <- as.data.frame(WS_results_ensembldb)
  res_WS_volcanoFULL <- res_WS_volcanoFULL %>%
    dplyr::filter(!is.na(log2FoldChange) & !is.na(padj))  
res_WS_volcanoFULL$Significant <- "NOT SIGNIFICANT"
### if log2Foldchange > 0.5 and padjvalue < 0.05, set as "UP" 
  res_WS_volcanoFULL$Significant[res_WS_volcanoFULL$log2FoldChange > 0.5 & res_WS_volcanoFULL$padj < 0.05] <- "UP"
### if log2Foldchange < -0.5 and padjvalue < 0.05, set as "DOWN"
  res_WS_volcanoFULL$Significant[res_WS_volcanoFULL$log2FoldChange < -0.5 & res_WS_volcanoFULL$padj < 0.05] <- "DOWN"
```
 
## Plot it:
```{r}
WSvolcanoFULL <- ggplot(res_WS_volcanoFULL, aes(x = log2FoldChange, y = -log10(padj), color = Significant)) +
    scale_colour_manual(values = c("#ff5c68", "black", "#4ada51")) +
    geom_point(alpha = 0.6, size = 1.5) +
    geom_hline(yintercept = -log10(0.05), linetype = "dashed", color = "blue") + 
    geom_vline(xintercept = c(-0.5, 0.5), linetype = "dashed", color = "blue") +
    geom_text_repel(data = subset(res_WS_volcanoFULL, Significant == c("UP", "DOWN")), max.overlaps = 20, aes(label = gene_name), vjust = 1.5, size = 3) +
    theme_minimal() +
    coord_fixed() +
    labs(x = "Log2 Fold Change", y = "-Log10 Adjusted P-value")
  theme(
    panel.border = element_blank(),
    axis.line = element_line(color = "black"))
    
```
```{r, echo=FALSE}
    plot(WSvolcanoFULL)
```
  
## **Aim 2: Venn diagram:**

## Make a subsetted WS that only has the significantly different genes:
```{r}
WS_filtered <- res_WS_volcanoFULL %>% 
    filter(Significant %in% c("UP", "DOWN"))
```
  
## Make a subsetted DE that only has the significantly different genes:
```{r}
DE_filtered <- res_DE_volcanoFULL %>% 
    filter(Significant %in% c("UP", "DOWN"))
```
  
## Merge the DE_filtered and WS_filtered to only include genes they were significantly changed in both conditions:
```{r}
filtered_DE_WS_merged <- merge(DE_filtered, WS_filtered, by = "gene_name") %>%
  dplyr::rename(Gene_label.DE = Gene_label.x, baseMean.DE = baseMean.x, log2FoldChange.DE = log2FoldChange.x, lfcSE.DE = lfcSE.x, stat.DE = stat.x, pvalue.DE = pvalue.x, padj.DE = padj.x, significant.DE = Significant.x) %>%
  dplyr::rename(Gene_label.WS = Gene_label.y, baseMean.WS = baseMean.y, log2FoldChange.WS = log2FoldChange.y, lfcSE.WS = lfcSE.y, stat.WS = stat.y, pvalue.WS = pvalue.y, padj.WS = padj.y, significant.WS = Significant.y)
```

## Combine the columns "gene_names" and "significant" from the significant WS and DE data frames:
```{r}
subsetted_WS_filtered <- select(WS_filtered, gene_name, Significant) %>%
    dplyr::rename(gene_names.WS = gene_name, significant.WS = Significant)
  
  subsetted_DE_filtered <- select(DE_filtered, gene_name, Significant) %>%
    dplyr::rename(gene_names.DE = gene_name, significant.DE = Significant)
```
  
## Make lists of up-regulated and down-regulated genes for WS and DE:
```{r}
upregulated_WS <- subsetted_WS_filtered %>%
    filter(significant.WS == "UP") %>%
    pull(gene_names.WS)
  
  downregulated_WS <- subsetted_WS_filtered %>%
    filter(significant.WS == "DOWN") %>%
    pull(gene_names.WS)
  
  upregulated_DE <- subsetted_DE_filtered %>%
    filter(significant.DE == "UP") %>%
    pull(gene_names.DE)
  
  downregulated_DE <- subsetted_DE_filtered %>%
    filter(significant.DE == "DOWN") %>%
    pull(gene_names.DE)
```
  
## Count overlaps for Venn diagram:
```{r}
venn_data <- list(
    WS = c(upregulated_WS, downregulated_WS),
    DE = c(upregulated_DE, downregulated_DE))
```
  
## Counts for up-regulated and down-regulated genes:
```{r}
ws_up_count <- length(upregulated_WS)
  ws_down_count <- length(downregulated_WS)
  de_up_count <- length(upregulated_DE)
  de_down_count <- length(downregulated_DE)
```
  
## Make the counts for the Venn diagram:
```{r}
venn_counts <- list(
    WS = paste("Up:", ws_up_count, "\nDown:", ws_down_count),
    DE = paste("Up:", de_up_count, "\nDown:", de_down_count))
```
  
## Plot the Venn diagram:
```{r}
venn_data_plot <- ggvenn(venn_data, 
         fill_color = c("#8d65b6", "#ff6cd0"), 
         show_percentage = FALSE, text_size = 7) +
    geom_text(data = data.frame(x = c(-0.8, 0.8), 
                                y = c(0.5, 0.5),
                                label = c(venn_counts$WS, venn_counts$DE)),
              aes(x = x, y = y, label = label), 
              size = 6, vjust = 1)
```

```{r, echo=FALSE}
plot(venn_data_plot)
```
 
## **Make A Table of the 10 shared genes:**
```{r}
DE_WS_merged_pretty <- filtered_DE_WS_merged %>%
      dplyr::select(Gene_label.DE, gene_name, log2FoldChange.DE, log2FoldChange.WS, padj.DE, padj.WS, significant.DE)
    
    colnames(DE_WS_merged_pretty) <- c("ENSEMBL Gene Name", "Hugo Gene Name", "log2FoldChange DE", "log2FoldChange WS", 
                                       "padj DE", "padj WS", "Significant")
```
    
## Make the table:
```{r}
DE_WS_pretty_table <-  DE_WS_merged_pretty %>%
      arrange(desc("log2FoldChange DE"))
    DE_WS_kable_table <- DE_WS_pretty_table 
    DE_WS_pretty_table %>%
      kable("html") %>%
      kable_styling(
        fixed_thead = T,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE) %>%
      row_spec(0, bold = TRUE, align = "c", color = "white", background = "#98A4D8") %>% 
      column_spec(7, background = ifelse(DE_WS_pretty_table$Significant == "DOWN", "#ffd5e9", "white")) 
```

## **Aim 2: Boxplots of the 10 shared significant genes:**

## Create the object using ensembl IDs:
```{r}
gene_list_boxplot <- c("ENSG00000245904", "ENSG00000259727", "ENSG00000107984", "ENSG00000129682",
                         "ENSG00000115602", "ENSG00000184408", "ENSG00000232931", "ENSG00000251562",
                         "ENSG00000166922", "ENSG00000261040")
  boxplot_norm_counts <- counts(dds, normalized = TRUE)[gene_list_boxplot, ]
  boxplot_data_long <- as.data.frame(t(boxplot_norm_counts)) %>%
    rownames_to_column("SampleID") %>%
    pivot_longer(cols = -SampleID, names_to = "ENSEMBL Gene", values_to = "Count")
```
  
## Add exposure column:
```{r}
colData_df <- as.data.frame(colData(dds))
  boxplot_data_long <- merge(boxplot_data_long, colData_df[, c("library.name", "Exposure")], 
                             by.x = "SampleID", by.y = "library.name", all.x = TRUE)
```
 
## Add Hugo Gene name column:
```{r}
boxplot_data_long <- merge(boxplot_data_long, DE_WS_pretty_table[, c("ENSEMBL Gene Name", "Hugo Gene Name")], 
                                  by.x = "ENSEMBL Gene", by.y = "ENSEMBL Gene Name", all.x = TRUE)
```

## Remove IC:
```{r}
boxplot_data_long_filtered <- boxplot_data_long %>%
     filter(Exposure != "IC")
```
 
## Plot it:
```{r}
box_all_plotted <- ggplot(boxplot_data_long_filtered, aes(x = Exposure, y = Count, fill = Exposure)) + 
   scale_fill_manual(values = c("FA" = "#d1b1fd", "DE" = "#88e4ff", "WS" = "#fbb8ff")) +
   geom_boxplot(outlier.shape = 16, outlier.colour = "#f3154e") + 
   labs(y = "Normalized Counts") + 
    theme_minimal() +
   facet_wrap(~ `Hugo Gene Name`, scales = "free_y", ncol = 3)+
   theme(strip.text = element_text(face = "bold"))
```

```{r, echo=FALSE}
   plot(box_all_plotted)
```
 
 
## **Aim 3: KEGG pathways:**

## DavidGene: DE significant, based on padj<0.05:
```{r}
DE_David <- as.data.frame(DE_results_ensembldb)
  DE_David <- DE_David %>%
      dplyr::filter(!is.na(log2FoldChange) & !is.na(padj)) 
  DE_David$Significant <- "NOT SIGNIFICANT" 
### if log2Foldchange > 0 and padjvalue < 0.05, set as "UP" 
   DE_David$Significant[DE_David$log2FoldChange > 0.0 & DE_David$padj < 0.05] <- "UP"
### if log2Foldchange < 0 and padjvalue < 0.05, set as "DOWN"
   DE_David$Significant[DE_David$log2FoldChange < 0.0 & DE_David$padj < 0.05] <- "DOWN"
```
    
## Upregulated only:
```{r}
DE_David_up <- DE_David %>%
      dplyr::filter(Significant == "UP")
    write.csv(DE_David_up$Gene_label, file = "DE_David_up", row.names = FALSE)
```
    
## Downregulated only:
```{r}
DE_David_down <- DE_David %>%
      dplyr::filter(Significant == "DOWN")
    write.csv(DE_David_down$Gene_label, file = "DE_David_down", row.names = FALSE)
```
    
    
## DavidGene: WS significant, based on padj<0.05:
```{r}
WS_David <- as.data.frame(WS_results_ensembldb)
    WS_David <- WS_David %>%
      dplyr::filter(!is.na(log2FoldChange) & !is.na(padj))  
    WS_David$Significant <- "NOT SIGNIFICANT"  
## if log2Foldchange > 0 and padjvalue < 0.05, set as "UP" 
    WS_David$Significant[WS_David$log2FoldChange > 0.0 & WS_David$padj < 0.05] <- "UP"
    ## if log2Foldchange < 0 and padjvalue < 0.05, set as "DOWN"
    WS_David$Significant[WS_David$log2FoldChange < 0.0 & WS_David$padj < 0.05] <- "DOWN"
## Up only: NOTE: no genes in this since none are upregulated
    WS_David_up <- WS_David %>%
      dplyr::filter(Significant == "UP")
    write.csv(WS_David_up$Gene_label, file = "WS_David_up", row.names = FALSE)
```
    
## Down only:
```{r}
WS_David_down <- WS_David %>%
      dplyr::filter(Significant == "DOWN")
    write.csv(WS_David_down$Gene_label, file = "WS_David_down", row.names = FALSE)  
```
    
## Making the bar plots for Kegg pathways:
### DE downregulated:
```{r}
DE_downregulated_Kegg <- data.frame(
      Pathway = c("Thermogenesis", "Oxidative phosphorylation", "Chemcial carcinogenesis (ROS)",
                  "Parkinson disease", "Diabetic cardiomyopathy", "Prion disease", "Amyotrophic lateral sclerosis", 
                  "Alzheimer disease", "Huntington Disease", "Pathways of neurodegeneration - multiple diseases",
                  "Retrograde endocannabinoid signaling", "Metabolic pathways", "Epstein-Barr virus infection"),
      Gene_Count = c("11", "9", "10", "10", "9", "10", "11", "11", "10", "11", "6", "16", "4"),
      P_value = c("0.00000013", "0.00000026", "0.0000011", "0.0000047", "0.0000053", "0.0000058", "0.0000083", "0.000013", 
                  "0.000014", "0.000081", "0.00073", "0.0051", "0.075")) 
      DE_downregulated_Kegg$P_value <- as.numeric(DE_downregulated_Kegg$P_value)
    DE_downregulated_Kegg$'-10logP' <- -10 * log10(DE_downregulated_Kegg$P_value)
    DE_downregulated_Kegg$Pathway <- reorder(DE_downregulated_Kegg$Pathway, DE_downregulated_Kegg$`-10logP`, FUN = max)
```
      
## Plot it:
```{r}
DE_downregulated_Kegg_bar <- ggplot(DE_downregulated_Kegg, aes(x = `-10logP`, y = Pathway)) +
      geom_bar(stat = "identity", fill = "#c0bcf9", color = "#9894d4", width = 0.7) +
      geom_text(aes(label = Gene_Count), size = 3,  hjust = -0.2, color = "#635db5") +
      theme_minimal() +
      labs(x = "-10 log P-value", y = "Pathway") +
      theme(axis.title.x = element_text(size = 12, color = "black"),  
            axis.title.y = element_text(size = 12, color = "black"), 
            panel.border = element_rect(color = "black", fill = NA),
            panel.grid.major = element_blank (),
            panel.grid.minor = element_blank())
```

```{r, echo=FALSE}
    plot(DE_downregulated_Kegg_bar)
```
    
    
## DE upregulated:
```{r}
DE_upregulated_Kegg <- data.frame(
        Pathway = c("Glycolysis", "Pentose phosphate pathway", "Biosynthesis of amino acids",
                    "HIF-1 signaling pathway", "Carbon metabolism", "Fructose and mannose metabolism", 
                    "Metabolic pathways", "Galactose metabolism", "Biosynthesis of cofactors", 
                    "Mineral absorption", "AMPK signaling pathway", "Renal cell carcinoma",
                    "Central carbon metabolism in cancer", "Biosynthesis of nucleotide sugars", "MAPK signaling pathway",
                    "Starch and sucrose metabolism", "Ferroptosis", "Porphyrin metabolism", "Amino acid sugar metabolism",
                    "Legionellosis", "Pathways in cancer", "Thyroid hormone signaling pathway", "Sphingolipid signaling pathway",
                    "Diabetic cardiomyopathy"),
        Gene_Count = c("9", "7", "9", "10", "10", "6", "28", "4", "6", "4", "5", "4", "4", "3", "7", "3", "3", "3", "3", "3", "9", "4", "4", "5"),
        P_value = c("0.000000089", "0.00000021", "0.00000022", "0.00000034", "0.00000058", "0.0000094", "0.00014", "0.0025", 
                    "0.010", "0.015", "0.021", "0.022", "0.023", "0.040", "0.044", "0.046", "0.050", "0.059", "0.069",
                    "0.083", "0.086", "0.087", "0.087", "0.1")) 
      DE_upregulated_Kegg$P_value <- as.numeric(DE_upregulated_Kegg$P_value)
      DE_upregulated_Kegg$'-10logP' <- -10 * log10(DE_upregulated_Kegg$P_value)
      DE_upregulated_Kegg$Pathway <- reorder(DE_upregulated_Kegg$Pathway, DE_upregulated_Kegg$`-10logP`, FUN = max)
```
    
## Plot it:
```{r}
DE_upregulated_Kegg_bar <- ggplot(DE_upregulated_Kegg, aes(x = `-10logP`, y = Pathway)) +
        geom_bar(stat = "identity", fill = "#c0bcf9", color = "#9894d4", width = 0.7) +
        geom_text(aes(label = Gene_Count), size = 3,  hjust = -0.2, color = "#635db5") +
        theme_minimal() +
        labs(x = "-10 log P-value", y = "Pathway") +
        theme(axis.title.x = element_text(size = 12, color = "black"),  
              axis.title.y = element_text(size = 12, color = "black"), 
              panel.border = element_rect(color = "black", fill = NA),
              panel.grid.major = element_blank (),
              panel.grid.minor = element_blank())
```

```{r, echo=FALSE}
    plot(DE_upregulated_Kegg_bar)
```
      
## WS downpregulated:
```{r}
WS_downregulated_Kegg <- data.frame(
        Pathway = c("Non-alcoholic fatty liver disease", "Oxidative phosphorylation", "Amyotrophic lateral sclerosis",
                    "Diabetic cardiomyopathy", "Alzheimer disease", "Chemical carcinogenesis (ROS)", 
                    "Cardiac muscle contraction", "Thermogenesis", "Parkinson disease", 
                    "Pathways of neurodegeneration - multiple diseases", "Prion disease", "Huntington disease",
                    "Metabolic pathways"),
        Gene_Count = c("6", "4", "5", "4", "5", "4", "3", "4", "4", "5", "4", "4", "8"),
        P_value = c("0.000034", "0.0046", "0.012", "0.014", "0.015", "0.018", "0.019", "0.020", 
                    "0.029", "0.029", "0.030", "0.040", "0.078")) 
      WS_downregulated_Kegg$P_value <- as.numeric(WS_downregulated_Kegg$P_value)
      WS_downregulated_Kegg$'-10logP' <- -10 * log10(WS_downregulated_Kegg$P_value)
      WS_downregulated_Kegg$Pathway <- reorder(WS_downregulated_Kegg$Pathway, WS_downregulated_Kegg$`-10logP`, FUN = max)
```
      
## Plot it:
```{r}
WS_downregulated_Kegg_bar <- ggplot(WS_downregulated_Kegg, aes(x = `-10logP`, y = Pathway)) +
        geom_bar(stat = "identity", fill = "#c0bcf9", color = "#9894d4", width = 0.7) +
        geom_text(aes(label = Gene_Count), size = 3,  hjust = -0.2, color = "#635db5") +
        theme_minimal() +
        labs(x = "-10 log P-value", y = "Pathway") +
        theme(axis.title.x = element_text(size = 12, color = "black"),  
              axis.title.y = element_text(size = 12, color = "black"), 
              panel.border = element_rect(color = "black", fill = NA),
              panel.grid.major = element_blank (),
              panel.grid.minor = element_blank())
```

```{r, echo=FALSE}
    plot(WS_downregulated_Kegg_bar)
```
     
      
## WS upregulated:
```{r}
WS_upregulated_Kegg <- data.frame(
        Pathway = c("RNA degradation"),
        Gene_Count = c("2"),
        P_value = c("0.077")) 
      WS_upregulated_Kegg$P_value <- as.numeric(WS_upregulated_Kegg$P_value)
      WS_upregulated_Kegg$'-10logP' <- -10 * log10(WS_upregulated_Kegg$P_value)
```
      
## Plot it:
```{r}
WS_upregulated_Kegg_bar <- ggplot(WS_upregulated_Kegg, aes(x = `-10logP`, y = Pathway)) +
        geom_bar(stat = "identity", fill = "#c0bcf9", color = "#9894d4", width = 0.7) +
        geom_text(aes(label = Gene_Count), size = 3,  hjust = -0.2, color = "#635db5") +
        theme_minimal() +
        labs(x = "-10 log P-value", y = "Pathway") +
        theme(axis.title.x = element_text(size = 12, color = "black"),  
              axis.title.y = element_text(size = 12, color = "black"), 
              panel.border = element_rect(color = "black", fill = NA),
              panel.grid.major = element_blank (),
              panel.grid.minor = element_blank())
```

```{r, echo=FALSE}
    plot(WS_upregulated_Kegg_bar)
```
     
## **Tables for Individual Pollutants:**

## Significant DE (parameters used in volcano and Venn)
### Rearrange the data a little bit:
```{r}
DE_filtered_for_table <- DE_filtered %>%
      select(Gene_label, gene_name, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj, Significant) %>%
      arrange(desc(Significant == "UP"))
    
    colnames(DE_filtered_for_table) <- c("ENSEMBL Gene Name", "Hugo Gene Name", "baseMean", 
                                         "log2FoldChange", "lfcSE", "stat", "pvalue", "padj", "Significant")
```
    
## Make the table:
```{r}
Significant_DE_table <-  DE_filtered_for_table %>%
      arrange(desc(log2FoldChange)) 
    DE_kable_table <- Significant_DE_table 
    Significant_DE_table %>%
      kable("html") %>%
      kable_styling(
        fixed_thead = T,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE) %>%
      row_spec(0, bold = TRUE, color = "white", background = "#98A4D8") %>% 
      column_spec(9, background = ifelse(Significant_DE_table$Significant == "UP", "#d6bde4", 
                                      ifelse(Significant_DE_table$Significant == "DOWN", "#ffd5e9", "white"))) %>%
      add_footnote("Log2 Fold Change parameters set to be greater than 0.5 for up-regulation and less than -0.5 for down-regulation. P-adjusted value set to less than 0.05 for both up and down-regulation.", notation = "symbol")
```
  
## Significant WS (parameters used in volcano and Venn)
### Rearrange the data a little bit:
```{r}
WS_filtered_for_table <- WS_filtered %>%
      select(Gene_label, gene_name, baseMean, log2FoldChange, lfcSE, stat, pvalue, padj, Significant) %>%
      arrange(desc(Significant == "UP"))
    
    colnames(WS_filtered_for_table) <- c("ENSEMBL Gene Name", "Hugo Gene Name", "baseMean", 
                                         "log2FoldChange", "lfcSE", "stat", "pvalue", "padj", "Significant")
```
    
    
## Make the table:
```{r}
Significant_WS_table <-  WS_filtered_for_table %>%
      arrange(desc(log2FoldChange)) 
    WS_kable_table <- Significant_WS_table 
    Significant_WS_table %>%
      kable("html") %>%
      kable_styling(
        fixed_thead = T,
        bootstrap_options = c("striped", "hover", "condensed"),
        full_width = FALSE) %>%
      row_spec(0, bold = TRUE, align = "c", color = "white", background = "#98A4D8") %>% 
      column_spec(9, background = ifelse(Significant_WS_table$Significant == "UP", "#d6bde4", 
                                         ifelse(Significant_WS_table$Significant == "DOWN", "#ffd5e9", "white"))) %>%
      add_footnote("Log2 Fold Change parameters set to be greater than 0.5 for up-regulation and less than -0.5 for down-regulation. P-adjusted value set to less than 0.05 for both up and down-regulation.", notation = "symbol")
```
    

    